How does an Experimental Software Engineering Lab Work? An Ethnographic Study

### 2023-07-25 tex/research-execution.tex
Last changes
Last changes  made before the submission.
Last changes
These are the last changes previous the submission

\section{Ethnographic study}\label{sec-reseach-execution}
We report the results of the ethnographic study in chronological order: First, we describe the analysis of the RGUS's preferred literature sources and the review of the RGUS' specific experimental material. Later, we comment on our experiences replicating an existing experiment. Finally, we describe how we formalized the information into concept and process models using interviews, focus groups, and workshops.

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support} \cite{Strandberg-2019-Ethical-Interviews-SE} \cite{Yang-2021-interview-study-developers}. Accordingly, the researchers started conducting interviews to clarify the RGUS' knowledge and convert it directly into models.

The interviews with RGUS members provided only public, limited knowledge about experimentation. The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}. We did study the literature.

However, communication became problematic when the interviews progressed into specialized areas, e.g., concrete families of experiments. For example, the experimenters supported their story-telling in specific experimental materials that the researchers did not have. Nor did the researchers have the knowledge required to understand the experimenters' discussions. At this point, the need for a different research methodology --ethnography-- became apparent. Consequently, the researchers switched minds and continued at a slower pace. We carried out additional activities to understand the standard literature and experimental material. The interaction with the experimenters also changed. For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS families of experiments.

\subsubsection{Analysis of the relevant literature}
The experimenters use the books by Wohlin et al. \cite{Wohlin-2000-Experimentation-SE}, and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process. They also utilize specific papers that advise how to experiment in SE, e.g., Basili et al. \cite{Basili-1986-ESE}, and Kitchenham et al. \cite{Kitchenham-2002-empirical-research-guidelines-SE}. Basili et al.'s work \cite{Basili-1999-families-experiments} was also considered relevant due to the experimenters' interest in carrying out experimental replications.

We reviewed the literature in-depth. We noticed that the sources do not use coherent terminology. At the outset, it seemed to the researchers that the literature describes different experimental procedures. They later verified that the different terms refer to essentially the same activities, e.g.:

\begin{itemize}
	\item Wohlin et al. \cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report".
	\item Juristo \& Moreno \cite{Juristo-2001-SE-experimentation}: "Objective Definition, Design, Execution and Analysis".
	\item Basili et al. \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation".
	\item Kitchenham et al. \cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data collection, Analysis, Presentation of results, and Interpretation of results".
\end{itemize}

We looked for additional ESE books to verify whether terminological diversity was a widespread problem. We only found a new edition of Wohlin et al.'s book \cite {Wohlin-2012-experimentatio-SE}, and the specialized books by Runeson et al. \cite{Runenson-2012-case-study-SE} and Kitchenham et al. \cite{Kitchenham-2015-Evidence-Based-SE}. These books share standard terms, as expected, but they use diverse terms throughout the text, contributing to the terminological diversity.

\subsubsection{Review of experimental material}\label{subsubsec-review-experimental-material}
The RGUS' experimental materials are a promising source of information since they represent the work performed during almost two decades of experimentation and reflect the tacit knowledge \cite{Polanyi-1996-tacit-k} present in the group. However, accessing the experimental materials was challenging; materials were scattered and disorganized. We needed the experimenters' help to find specific documents/data files, understand their structure and contents, and gain further access to other related materials.

The experimental materials evolve. Experiments vary widely in documentary support, independently of their date. The experimental family explains the differences better than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.

Concerning the experimental process, all experimental materials describe the activities carried out in a general way. The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge. Terminological diversity also affects experimental materials, especially among families of experiments.

@@

\subsubsection{Experiment selection}
Instead of designing an experiment from scratch, the researchers opted to replicate an experiment previously conducted by the RGUS. The replication has been published elsewhere.

Conducting a replication of the experimenters' previous work attracted their attention and guaranteed collaboration. Furthermore, the replication would give us the chance to recreate the design, execution, analysis, and reporting of an experiment that, until then, we only knew due to the examination of the written materials. Finally, the replication would help us better understand the experimenters' motivations and decisions.

\subsubsection{Replication planning}
Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE} \cite{Vegas-2006-communication-researchers} \cite{Miller-2005-replicating-SE-experiments} \cite{Gomez-2014-understanding-replication} \cite{Demagalhaes-2015-replications-SE} \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities. For example, how to approach the experimental design, which experimental objects are necessary, and in which order. We requested the cooperation of the experimenters who ran the baseline experiment to schedule the experimental activities and adapt the design to the new context where the replication would be run.

@@

The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined in advance all relevant aspects. The measurement was the most complex and controversial aspect. Let's say that the experimental subjects return test cases, but these test cases are manually interpreted and not run. The experimenters' cooperation was again needed to complete the measurement. Even if the measurement procedure looked controversial, it made complete sense after discussion. We learned that the experimental protocols depend on the experimental task, which in turn depends on the research area. The remaining steps, e.g., calculating measurements from raw data and formatting, did not imply any uncertainty. It was possible to reproduce the duplicate files used in the baseline experiment.

\subsubsection{Analysis and reporting}
We sought help to analyze and report the replication since we did not have statistical training. In retrospect, both the analysis and the report were simple to perform since the experimental design used (\textit{cross-over}) was associated with a well-defined analysis method \cite{Vegas-2016-crossover-designs-experiments-SE}. On the other hand, there were reporting guidelines available \cite{Carver-2010-guidelines-replication-SE}. The experimenters confirmed that, generally, the analysis and reporting of experiments do not imply substantial SE challenges.

The differences observed between replication and the baseline experiment confirmed our belief that, at least in the case of the RGUS, there are no clear guidelines that allow literal, external, and independent replication \cite{Gomez-2014-understanding-replication} of the original experiment, irrespective of the existence of incidents in the replication's execution. As a corollary, the original experimenters' presence was necessary to make the baseline experiment results and the replication comparable.

@@

\subsubsection{Semi-structured interviews}
We decided to organize new series of semi-structured interviews with the experimenters. The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on the replication of \textit{specific families of experiments}. Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.

We created an open-ended questionnaire to motivate the experimenters to give a spontaneous but relatively ordered account of their experiences. We gave up note-taking and captured the interviews on video, including the experimenters' body language. We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative. We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols}. We created a report after each interview, e.g., see this \href{https://zenodo.org/record/7102137#.YytgDOzMLUI}{\ul{link}}.

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Boseto-Proccess}
	\caption{Example of a sketch made by experimenters.}
	\label{fig-proceso-exp-boseto}
\end{figure}

@@

\begin{itemize}
	\item \textit{Preliminary} experimental process model: We realized that the way the RGUS conducts experiments resembles the processes we are used to in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}. In the end, they all are processes. Hence, we abstracted six high-level experimental processes: support, generation of pieces of knowledge, publications, synthesis, basic, and organization. Each process has several related activities. For example, the \textit{Basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).

@@

An experimenter does not usually carry out all activities, although they know how to perform them (maybe not flawlessly). In turn, different responsibilities are typically assigned to the same experimenter. We identified the following roles (the names suggest their responsibilities): (1) \textit{Research Manager}, (2) \textit{Experiment Manager} and (3) \textit{Senior Experimenter}. These roles are responsible for: (1) planning the experimental research, (2) managing the logistics, and (3) conducting the experiment in practice, respectively. The study showed overlapping responsibilities among roles, e.g., the domain and topic selection for the experiment.

@@

As we said above, we organized the semi-structured interviews around families of experiments. It was a fortunate mistake. We found out that the families of experiments are pretty homogeneous. The experimenters involved in a family quickly agree and provide a coherent picture of the experimental process. However, experimenters from different families disagree on many aspects, particularly the specific arrangements used in the experimental families. For instance, the experimenters never agreed on the activities related to the experimental design.

We aimed to solve the disagreements using a consensus-reaching technique, e.g., DELPHI \cite{Dalkey-1967-Delphi}. However, social pressure was not a problem. We decided to organize focus groups. The most senior experimenter led the meetings as a moderator, but only to keep the discussion on track. The researchers only observed. Each session lasted at least two hours. Before the focus groups, the experimenters reviewed the concept models. We started with the concept models because they were more straightforward and smaller. The process models/workflow were later updated based on the changes to the concept model.

The focus groups prompted a substantial growth in the models' details after heated discussions that re-confirmed the existence of terminological diversity \textit{even among experimenters who carry out similar activities within the same family of experiments}. The most likely reason seems to be the training that each experimenter received. Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions. The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches} \cite{montgomery-2019-Design-Analysis-Experiments} did not drive self-education. Few of these materials existed when the experimenters started their careers. The experimenters chose the training materials according to their preferences and the experimental families in which they participated. Terminological diversity was the logical consequence.

The experimenters did not reach a consensus during the focus group exercise. The discussions with experimenters of different families exhibited a characteristic circular pattern: An experimenter triggered a change that, later, was undone by another. Substantial agreement was only possible \textit{within each family of experiments}. The progress we achieved during the replication phase of the ethnographic research (see Section~\ref {replication}) was possible due to our focus on concrete families of experiments, a fact that we completely ignored at that time.

@@

\subsubsection{Role-focused workshops}\label{subsubsec-focus-groups-role}
We did not aim to explore SE family-specific experimental processes. Albeit enjoyable, we perceived it as future research. In turn, we tried to take advantage of the limited PhD time available to improve the models by switching the perspective from experimental families to \textit{experimental roles}. There was no reason to think that roles could contribute to the agreement. However, we used the concept model to drive the focus groups, not the process models. Roles are connected with activities, not concepts. Activities looked more uniform than concepts. It was worth trying.

@@

We could also break down the conceptual model into role-specific conceptual models. They represent groups of concepts used by the experimenters when they play a specific role. Most experimental activities are role-specific, e.g., the senior experimenter role uses the run-kit concept exclusively. The role-specific conceptual models are available through these links: \href{https://zenodo.org/record/7102431#.YyuFvuzMLUI}{\ul{1}}, \href{https://zenodo.org/record/7102450#.YyuG2OzMLUI}{\ul{2}}, \href{https://zenodo.org/record/7102464#.YyuIkOzMLUI}{\ul{3}}.

>>>>

\section{Ethnographic study}\label{sec-reseach-execution}
We report the results of the ethnographic study chronologically: First, we describe the analysis of the RGUS's preferred literature sources and review the RGUS' specific experimental material. Later, we comment on our experiences replicating an existing experiment. Finally, we describe how we formalized the information into concept and process models using interviews, focus groups, and workshops.

\subsection{First Approximation}
In the first stages of the investigation, the researchers did not have an ethnographic methodology in mind but something less demanding, in line with existing qualitative studies, e.g., \cite{Johanssen-2019-Continuous-SE-support}, \cite{Strandberg-2019-Ethical-Interviews-SE}, and \cite{Yang-2021-interview-study-developers}. Accordingly, the researchers conducted interviews to clarify the RGUS' knowledge and convert it directly into models.

The interviews with RGUS members provided only public, limited knowledge about experimentation. The experimenters believed that interviews were unnecessary because the researchers could acquire knowledge by studying the literature that the experimenters considered the standard, e.g., \cite{Wohlin-2000-Experimentation-SE}. Hence, we did study the literature too.

However, communication became problematic when the interviews progressed into specialized areas, e.g., concrete families of experiments. For example, the experimenters supported their storytelling using specific experimental materials the researchers did not have. Nor did the researchers know to understand the experimenters' discussions. At this point, the need for a different research methodology --ethnography-- became apparent. Consequently, the researchers switched minds and continued at a slower pace. We carried out additional activities to understand the standard literature and experimental material. The interaction with the experimenters also changed. For example, the researchers participated first as experimental subjects and later as team members in one of the RGUS' families of experiments.

\subsubsection{Analysis of the relevant literature}
The experimenters use the books by Wohlin et al. \cite{Wohlin-2000-Experimentation-SE} and Juristo \& Moreno \cite{Juristo-2001-SE-experimentation} as primary sources for their experimentation process. They also utilize specific papers that advise how to experiment in SE, e.g., Basili et al. \cite{Basili-1986-ESE}, and Kitchenham et al. \cite{Kitchenham-2002-empirical-research-guidelines-SE}. Basili et al.'s work \cite{Basili-1999-families-experiments} was also considered relevant due to the experimenters' interest in carrying out experimental replications.

We reviewed the literature in depth. We noticed that the sources do not use coherent terminology. At the outset, it seemed to the researchers that the literature describes different experimental procedures. They later verified that the different terms refer to essentially the same activities, e.g.:

\begin{itemize}
	\item Wohlin et al. \cite{Wohlin-2000-Experimentation-SE}: "Experiment idea, Experiment scoping, Experiment planning, Experiment operation, Analysis \& interpretation, Presentation \& package, and Experiment report."
	\item Juristo \& Moreno \cite{Juristo-2001-SE-experimentation}: "Objective Definition, Design, Execution, and Analysis."
	\item Basili et al. \cite{Basili-1986-ESE}: "Definition, Planning, Operation, and Interpretation."
	\item Kitchenham et al. \cite{Kitchenham-2002-empirical-research-guidelines-SE}: "Experimental context, Experimental design, Conduct of the experiment and Data Collection, Analysis, Presentation of results, and Interpretation of results."
\end{itemize}

We looked for additional ESE books to verify whether terminological diversity was a widespread problem. We only found a new edition of Wohlin et al.'s book \cite {Wohlin-2012-experimentatio-SE}, and the specialized books by Runeson et al. \cite{Runenson-2012-case-study-SE} and Kitchenham et al. \cite{Kitchenham-2015-Evidence-Based-SE}. As expected, these books share standard terms but use diverse terms throughout the text, contributing to the terminological diversity.

\subsubsection{Review of experimental material}\label{subsubsec-review-experimental-material}
The RGUS' experimental materials are a promising source of information since they represent the work performed during almost two decades of experimentation and reflect the tacit knowledge \cite{Polanyi-1996-tacit-k} present in the group. However, accessing the experimental materials was challenging; materials were scattered and disorganized. We needed the experimenters' help to find specific documents/data files, understand their structure and contents, and gain further access to related materials.

The experimental materials evolve. Experiments vary widely in documentary support, independently of their date. The experimental family explains the differences more significantly than the time; the materials related to different families of experiments show significant differences, e.g., digital raw data versus physical papers written with raw data; or automatic tools to measure variables versus manual tools.

Concerning the experimental process, all experimental materials describe the activities carried out generally. The execution order and the relationship between experimental activities are unclear; they rely on the experimenters' tacit knowledge. Terminological diversity also affects experimental materials, especially among families of experiments.

@@

\subsubsection{Experiment selection}
Instead of designing an experiment from scratch, the researchers replicated an experiment previously conducted by the RGUS. The replication has been published elsewhere.

Conducting a replication of the experimenters' previous work attracted their attention and guaranteed collaboration. Furthermore, the replication would allow us to recreate the design, execution, analysis, and reporting of an experiment that, until then, we only knew due to the examination of the written materials. Finally, the replication would help us better understand the experimenters' motivations and decisions.

\subsubsection{Replication planning}
Various problems hinder replication \cite{Gallardo-2012-CG-PL-SE}, \cite{Vegas-2006-communication-researchers}, \cite{Miller-2005-replicating-SE-experiments}, \cite{Gomez-2014-understanding-replication}, \cite{Demagalhaes-2015-replications-SE}, \cite{Carver-2010-guidelines-replication-SE}, being the main one the absence of a detailed account of the experimental activities. For example, how to approach the experimental design, which experimental objects are necessary, and in which order. We requested the cooperation of the experimenters who ran the baseline experiment to schedule the experimental activities and adapt the design to the new context where the replication would be run.

@@

The existence of explicit experimental materials such as questionnaires, task descriptions, and programs significantly eased experiment conduction since the researchers defined in advance all relevant aspects. The measurement was the most complex and controversial aspect. Let us say that the experimental subjects return test cases, but these test cases are manually interpreted and not run. The experimenters' cooperation was again needed to complete the measurement. Even if the measurement procedure looked controversial, it made complete sense after discussion. We learned that the experimental protocols depend on the experimental task, which depends on the research area. The remaining steps, e.g., calculating measurements from raw data and formatting, did not imply any uncertainty. It was possible to reproduce the duplicate files used in the baseline experiment.

\subsubsection{Analysis and reporting}
We sought help to analyze and report the replication since we did not have statistical training. In retrospect, both the analysis and the report were simple to perform since the experimental design used (\textit{cross-over}) was associated with a well-defined analysis method \cite{Vegas-2016-crossover-designs-experiments-SE}. On the other hand, there were reporting guidelines available \cite{Carver-2010-guidelines-replication-SE}. The experimenters confirmed that the analysis and reporting of experiments generally do not imply substantial SE challenges.

The differences observed between replication and the baseline experiment confirmed our belief that, at least in the case of the RGUS, there are no clear guidelines that allow literal, external, and independent replication \cite{Gomez-2014-understanding-replication} of the original experiment, irrespective of the existence of incidents in the replication's execution. As a corollary, the original experimenters' presence was necessary to compare the baseline experiment results and replication.

@@

\subsubsection{Semi-structured interviews}
We decided to organize a new series of semistructured interviews with the experimenters. The key difference with the first approximation was that we would not discuss the SE experimental process \textit{in general}, but we would focus on replicating \textit{specific families of experiments}. Given the experience acquired during the replication, we believed this strategy could expose the tacit experimenters' knowledge.

We created an open-ended questionnaire to motivate the experimenters to give a spontaneous but relatively ordered account of their experiences. We gave up note-taking and captured the interviews with video, including the experimenters' body language. We also collected paper sketches (e.g., see Fig.~\ref{fig-proceso-exp-boseto}) that some experimenters used to support their narrative. We analyzed videos and sketches using protocol analysis \cite{Pressley-1995-verbal-protocols}. We created a report after each interview, e.g., see this \href{https://zenodo.org/record/7102137#.YytgDOzMLUI}{\ul{link}}.

\begin{figure}[htbp!]
	\centering
	\includegraphics[width=\columnwidth]{images/Boseto-Proccess}
	\caption{example of a sketch made by experimenters.}
	\label{fig-proceso-exp-boseto}
\end{figure}

@@

\begin{itemize}
	\item \textit{Preliminary} experimental process model: We realized that the RGUS conducts experiments that resemble the processes we use in SE, e.g., ISO/IEC 12207 \cite{ISO-IEC-IEEE-12207}. In the end, they all are processes. Hence, we abstracted six high-level experimental processes: support, generation of pieces of knowledge, publications, synthesis, basic, and organization. Each process has several related activities. For example, the \textit{basic process} includes the activities: pre-execution (context adaptation, experiment redesign, and experiment design), execution (pre-session, in-session, and post-session), and post-execution (raw data generation, data processing, data analysis, and data interpretation) (see this \href{https://zenodo.org/record/7102301#.Yyt0GOzMLUI}{\ul{link}}).

@@

An experimenter does not usually carry out all activities, although they know how to perform them (maybe not flawlessly). In turn, different responsibilities are typically assigned to the same experimenter. We identified the following roles (the names suggest their responsibilities): (1) \textit{Research Manager}, (2) \textit{Experiment Manager}, and (3) \textit{Senior Experimenter}. These roles are responsible for (1) planning the experimental research, (2) managing the logistics, and (3) conducting the experiment in practice, respectively. The study showed overlapping responsibilities among roles, e.g., the domain and topic selection for the experiment.

@@

As we said above, we organized the semistructured interviews around families of experiments. It was a fortunate mistake. We found out that the families of experiments are pretty homogeneous. The experimenters involved in the family quickly agree and provide a coherent picture of the experimental process. However, experimenters from different families disagree on many aspects, particularly the specific arrangements used in the experimental families. For instance, the experimenters never agreed on the activities related to the experimental design.

We aimed to solve the disagreements using a consensus-reaching technique, e.g., DELPHI \cite{Dalkey-1967-Delphi}. However, social pressure was not a problem. We decided to organize focus groups. The most senior experimenter led the meetings as a moderator, but only to keep the discussion on track. The researchers only observed. Each session lasted at least two hours. Before the focus group, the experimenters reviewed the concept models. We started with the concept models because they were more straightforward and smaller. The process models/workflow were later updated based on the changes to the concept model.

The focus groups prompted a substantial growth in the models' details after heated discussions that reconfirmed the existence of terminological diversity \textit{even among experimenters who carried out similar activities within the same family of experiments}. The most likely reason seems to be the training that each experimenter received. Talking about "training" in the context of the RGUS is probably incorrect since its members are self-educated, with few exceptions. The preferred SE experimental literature, e.g., \cite{Creswell-2009-Method-Approaches}, \cite{montgomery-2019-Design-Analysis-Experiments}, did not drive self-education. Few of these materials existed when the experimenters started their careers. The experimenters chose the training materials according to their preferences and the experimental families in which they participated. Terminological diversity was the logical consequence.

The experimenters did not reach a consensus during the focus group exercise. The discussions with experimenters of different families exhibited a characteristic circular pattern: An experimenter triggered a change later undone by another. Substantial agreement was only possible \textit{within each family of experiments}. The progress we achieved during the replication phase of the ethnographic research (see Section~\ref {replication}) was possible due to our focus on concrete families of experiments, a fact that we completely ignored at that time.

@@

\subsubsection{Role-focused workshops}\label{subsubsec-focus-groups-role}
We did not aim to explore SE family-specific experimental processes. Albeit enjoyable, we perceived it as future research. We tried to use the limited Ph.D. time to improve the models by switching the perspective from experimental family to \textit{experimental role}. There was no reason to think that roles could contribute to the agreement. However, we used the concept model to drive the focus groups, not the process models. Roles are connected to activities, not concepts. Activities looked more uniform than concepts. It was worth trying.

@@

We could also break down the conceptual model into role-specific conceptual models. They represent groups of concepts the experimenters use when they play a specific role. Most experimental activities are role-specific, e.g., the senior experimenter role uses the run-kit concept exclusively. The role-specific conceptual models are available through these links: \href{https://zenodo.org/record/7102431#.YyuFvuzMLUI}{\ul{1}}, \href{https://zenodo.org/record/7102450#.YyuG2OzMLUI}{\ul{2}}, \href{https://zenodo.org/record/7102464#.YyuIkOzMLUI}{\ul{3}}.

### 2023-07-25 tex/research-method.tex
Last changes
These are the last changes previous the submission

Ethnography identifies details that are ignored with other methods. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted for the study of a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative, i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.

We followed Runenson et al.'s case studies guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:

@@

To avoid confusion, from this point on we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters". We will talk about ourselves the authors as the "the researchers". We the authors/researchers play two different roles in the investigation:

\begin{itemize}
\item \textbf{Efra\'in R. Fonseca C.} and Edison G. Espinosa conducted the ethnographic study as part of their Ph.D. Edison G. Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of this Ph.D. This is the reason why he is not authoring this paper. The personal opinions poured in Section~\ref{sec-reseach-execution} belong to these two researchers, particularly Efra\'in R. Fonseca C.
\item \textbf{Oscar Dieste} and \textbf{Natalia Juristo} supervised the investigation. Their views and opinions are not reflected in this manuscript.
\end{itemize}

@@

Ethnography information gathering is characterized by being cyclical, iterative, and incremental. The data acquisition process was based on a taxonomy of information gathering techniques, categorized according to the level of contact with the primary source of information. Such taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}. Table \ref{tbl-tecnica-fuente} describes the information-gathering techniques and instruments and the corresponding information sources to which they can be applied.

@@

The order in which researchers contact information sources is essentially random. The frequency of contact depends on the researchers' needs to answer research questions during the ethnographic research and the availability of each source.

@@

\begin{itemize}
\item \textbf{Preparation of materials}: The information sources are classified and categorized so experimenters and researchers can manage them.
\item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
\item \textbf{Verification of information}: This is probably the most complex activity. The knowledge obtained from the sources complements each other. We will use triangulation to explain the findings and identify new areas of inquiry. All findings are discussed with the experimenters to avoid misunderstandings and verify that the resulting activities are those that they carry out in practice.
\item \textbf{Generation of results}: Finally, we obtained the results as final products after several iterations of comparison, verification, and consensus between researchers and experimenters. The expected results will be models (conceptual and process) and the practices derived from the interviews with the experimenters.
\end{itemize}

>>>>

Ethnography identifies details that are ignored by or in other methods. The disadvantage is the time required for learning \cite{Sharp-2016-Ethnographic-Studies-ESE}. Thus, we opted for the study of a single experimental research group over a long period instead of working halfway with several experimental research groups over shorter periods. This decision was also motivated by the challenge of finding several open-minded experimental research groups willing to participate in this research. We selected a representative i.e., respected and well-known, research group to guarantee (to the extent possible) the reliability of the results, their validity, and the findings' scientific significance \cite{sjoberg-2007-future-empical-methods}.

We followed Runenson et al.'s case study guidelines \cite{Runenson-2012-case-study-SE} wherever they apply to ethnographical studies. Our research protocol is organized along three axes:

@@

To avoid confusion, from this point on we will refer to the "ESE researchers" with whom we interact during the investigation as "the experimenters". We will talk about ourselves the authors as "the researchers". We authors/researchers play two different roles in the investigation:

\begin{itemize}
\item \textbf{Efra\'in R. Fonseca C.} and Edison G. Espinosa conducted the ethnographic study as part of their Ph.D. Edison G. Espinosa abandoned the ethnographic study early (during the investigation reported in Section~\ref{replication}) and changed the orientation of this Ph.D. This is the reason he is not authoring this paper. The personal opinions poured in Section~\ref{sec-reseach-execution} belong to these two researchers, particularly Efra\'in R. Fonseca C.
\item \textbf{Oscar Dieste} and \textbf{Natalia Juristo} supervised the investigation. Their views and opinions are not reflected in this manuscript.
\end{itemize}

@@

Ethnography information gathering is characterized by being cyclical, iterative, and incremental. The data acquisition process was based on a taxonomy of information gathering techniques, categorized according to the level of contact with the primary source of information. Such a taxonomy proposes three levels of information gathering: direct participation with the source (level 1), indirect participation with the source (level 2), and study of the experiment material without the participation of the source (level 3) \cite{Lethbridge-2005-data-collection-techniques-SE}. Table \ref{tbl-tecnica-fuente} describes the information gathering techniques and instruments and the corresponding information sources to which they can be applied.

@@

The order in which researchers contact information sources is essentially random. The frequency of contact depends on the researchers' need to answer research questions during the ethnographic research and the availability of each source.

@@

\begin{itemize}
\item \textbf{Preparation of materials}: The information sources are classified and categorized, so experimenters and researchers can manage them.
\item \textbf{Information acquisition}: While analyzing the information sources, researchers gradually acquire basic knowledge about experimentation, which helps them understand the experimenters' knowledge and behavior more easily.
\item \textbf{Verification of information}: This is probably the most complex activity. The knowledge obtained from the sources complements each other. We will use triangulation to explain the findings and identify new areas of inquiry. All findings are discussed with the experimenters to avoid misunderstandings and verify that the resulting activities are those that they carry out in practice.
\item \textbf{Generation of results}: Finally, we obtained the results as final products after several iterations of comparison, verification, and consensus between researchers and experimenters. The expected results will be models (conceptual and process) and practices derived from the interviews with the experimenters.
\end{itemize}
